---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# immunedeconv2

<!-- badges: start -->
<!-- badges: end -->

The goal of immunedeconv2 is to ...

## Installation

You can install the released version of immunedeconv2 from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("immunedeconv2")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("PelzKo/immunedeconv2")
```
## Example

Binding two factors via `fbind()`:

```{r}
library(immunedeconv2)
a <- factor(c("character", "hits", "your", "eyeballs"))
b <- factor(c("but", "integer", "where it", "counts"))
```

Simply catenating two factors leads to a result that most don't expect.

```{r}
c(a, b)
```

The `fbind()` function glues two factors together and returns factor.

```{r}
fbind(a, b)
```

Often we want a table of frequencies for the levels of a factor. The base `table()` function returns an object of class `table`, which can be inconvenient for downstream work.

```{r}
set.seed(1234)
x <- factor(sample(letters[1:5], size = 100, replace = TRUE))
table(x)
```

The `fcount()` function returns a frequency table as a tibble with a column of factor levels and another of frequencies:

```{r}
fcount(x)
```
